---
title: "Ordination with vegan"
author: "Naupaka Zimmerman"
date: "March 3, 2016"
output:
  ioslides_presentation:
    highlight: tango
    keep_md: yes
    widescreen: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "", cache = TRUE, crop = TRUE)
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)
```



# Ordination plots



## Unconstrained ordination

**What does it mean?**

* First we look for major variation, then try to attribute it to environmental variation
* vs. constrained ordination, where we only want to see what can be explained by environmental variables of interest



## Unconstrained ordination

**What does it mean?**

* First we look for major variation, then try to attribute it to environmental variation
* vs. constrained ordination, where we only want to see what can be explained by environmental variables of interest

**Examples of unconstrained ordination:**

* Correspondance Analysis - CA
* Principal Components Analysis - PCA
* Principal Coordinates Analysis - PCoA
* Nonmetric Multidimensional Scaling - NMDS



## Unconstrained ordination

**What does it mean?**

* First we look for major variation, then try to attribute it to environmental variation
* vs. constrained ordination, where we only want to see what can be explained by environmental variables of interest

**Examples of unconstrained ordination:**

* Correspondance Analysis - CA
* Principal Components Analysis - PCA
* Principal Coordinates Analysis - PCoA
* **Nonmetric Multidimensional Scaling - NMDS**



## Before we get started

Housekeeping
```{r housekeeping, eval=FALSE}
setwd("your/working/dir")
```

```{r loading-1, results='hide'}
library("vegan")
MLM.otus <- read.csv("data/MLM_data_otus.csv", row.names = 1, header = TRUE)
MLM.env <- read.csv("data/MLM_data_env.csv", row.names = 1, header = TRUE)
```

## Before we get started | OTUs

```{r loading-2}
dim(MLM.otus)
head(MLM.otus[,1:10], n=1)
```



## Before we get started | environment

```{r loading-3}
head(MLM.env, n=1)
```

## Before we get started | environment

```{r loading-3.5}
summary(MLM.env)
```



## Basic ordination and plotting

There are two basic NMDS ordination functions:

* `isoMDS()` from the MASS package
* `monoMDS()` in vegan

Vegan also has a wrapper function for doing NMDS ordinations using best practices:

* `metaMDS()`

This will do handy things like try to standardize your data if necessary and perform rotation, among other things.



## Basic ordination and plotting

```{r NMDS-1, results='hide'}
MLM.bray.ord <- metaMDS(MLM.otus, distance = "bray", k = 2, trymax = 50)
```

**Show in RStudio**



## Basic ordination and plotting (using all defaults)

```{r NMDS-2, fig.height=5}
plot(MLM.bray.ord)
```




## Basic ordination and plotting (just plots)

```{r NMDS-3, fig.height=5}
plot(MLM.bray.ord, display = "sites")
```



## Basic ordination and plotting (just species)

```{r NMDS-4, fig.height=5}
plot(MLM.bray.ord, display = "species")
```



## Site names instead of points

```{r NMDS-5, fig.height=5}
plot(MLM.bray.ord, display = "sites", type = "t")
```



## Site names instead of points

```{r NMDS-5.2, fig.height=5}
plot(MLM.bray.ord, display = "sites")
set.seed(314) ## make reproducible
ordipointlabel(MLM.bray.ord, display = "sites", scaling = 3, add = TRUE)
```


## Larger points

```{r NMDS-6, fig.height=5}
plot(MLM.bray.ord, display = "sites", cex=2)
```


## Modifying the display of the points with environmental data 

* Color
* Shape
* Size



## Modifying the color of points {.smaller}
```{r NMDS-7, fig.height=4.5, crop = TRUE}
colors.vec <- c("red", "blue")
plot(MLM.bray.ord, display = "sites", type = "n")
points(MLM.bray.ord, display = "sites", cex=2, pch = 21, 
       col = colors.vec[MLM.env$side_of_island], 
       bg = colors.vec[MLM.env$side_of_island])
legend("topright", legend = levels(MLM.env$side_of_island), bty = "n",
                      col = colors.vec, pch = 21, pt.bg = colors.vec)
```



## Modifying the shape of points | pch()
```{r NMDS-8, fig.width=11, echo=FALSE}
par(mar = rep(0,4))
plot(c(-1, 26), 0:1, type = "n", axes = FALSE)
text(0:25, 0.4, 0:25, cex = 1.5)
points(0:25, rep(0.3, 26), pch = 0:25, bg = "grey", cex = 2)
```



## Modifying the shape of points {.smaller}
```{r NMDS-9, fig.height=4.5}
shapes.vec <- c(21, 22)
plot(MLM.bray.ord, display = "sites", type = "n")
points(MLM.bray.ord, display = "sites", cex=2, bg = "black", 
       pch = shapes.vec[MLM.env$flow_type])
legend("topright", legend = levels(MLM.env$flow_type), bty = "n",
                      col = "black", pch = shapes.vec, pt.bg = "black")
```



## Modifying the shape and color of points {.smaller}
```{r NMDS-10, fig.height=4}
colors.vec <- c("red", "blue")
shapes.vec <- c(21, 22)
plot(MLM.bray.ord, display = "sites", type = "n")
points(MLM.bray.ord, display = "sites", cex=2, 
       pch = shapes.vec[MLM.env$flow_type], 
       col = colors.vec[MLM.env$side_of_island], 
       bg = colors.vec[MLM.env$side_of_island])
legend("topright", legend = levels(MLM.env$side_of_island), bty = "n",
                      col = colors.vec, pch = 21, pt.bg = colors.vec)
legend(1.4,1.05, legend = levels(MLM.env$flow_type), bty = "n",
                      col = "black", pch = shapes.vec, pt.bg = "black")
```



## Adding other layers

```{r NMDS-11, fig.height=5}
# Just points
plot(MLM.bray.ord, display = "sites", cex=2)
```



## Adding other layers

```{r NMDS-12, fig.height=5}
plot(MLM.bray.ord, display = "sites", cex=2)
ordihull(MLM.bray.ord, groups = MLM.env$site_ID, label = FALSE)
```



## Adding other layers

```{r NMDS-13, fig.height=5}
plot(MLM.bray.ord, display = "sites", cex=2)
ordihull(MLM.bray.ord, groups = MLM.env$site_ID, label = FALSE, col = "blue")
```



## Adding other layers

```{r NMDS-14, fig.height=5}
plot(MLM.bray.ord, display = "sites", cex=2)
ordihull(MLM.bray.ord,groups = MLM.env$site_ID, label = FALSE, col = "blue")
ordispider(MLM.bray.ord,groups = MLM.env$site_ID, label = TRUE)
```



## Adding other layers

```{r NMDS-15, fig.height=5}
# Plot first, then add layers
plot(MLM.bray.ord, display = "sites", cex=2)
```



## Adding other layers

```{r NMDS-16, fig.height=5}
plot(MLM.bray.ord, display = "sites", cex=2)
ordispider(MLM.bray.ord,groups = MLM.env$site_ID, label = TRUE)
```



## Adding other layers - axes scaling

```{r NMDS-17, fig.height=5}
plot(MLM.bray.ord, type = "n")
```



## Adding other layers - axes scaling

```{r NMDS-18, fig.height=5}
plot(MLM.bray.ord, type = "n")
points(MLM.bray.ord,display = "sites", cex = 2)
```



## Adding other layers - axes scaling

```{r NMDS-19, fig.height=5}
plot(MLM.bray.ord, display = "sites", type = "n")
```



## Adding other layers - axes scaling

```{r NMDS-20, fig.height=5}
plot(MLM.bray.ord, display = "sites", type = "n")
points(MLM.bray.ord, display = "sites", cex = 2)
```



## Adding other layers

```{r NMDS-21, fig.height=5}
plot(MLM.bray.ord, display = "sites", type = "n")
points(MLM.bray.ord,display = "sites", cex = 2)
ordispider(MLM.bray.ord,groups = MLM.env$site_ID, label = TRUE)
```



## Adding other layers

```{r NMDS-22, fig.height=5}
plot(MLM.bray.ord, display = "sites", type = "n")
points(MLM.bray.ord, display = "sites", cex = 2)
ordiellipse(MLM.bray.ord,groups = MLM.env$site_ID, label = FALSE)
```



## Adding other layers
```{r NMDS-23, fig.height=5, message=FALSE, results='hide'}
plot(MLM.bray.ord, display = "sites", type = "n")
points(MLM.bray.ord,display = "sites", cex = 2)
ordisurf(MLM.bray.ord,MLM.env$elevation_m, add = TRUE)
```



## Vectors in ordination space

```{r NMDS-24}
MLM.bray.ord.elev.fit <- envfit(MLM.bray.ord ~ elevation_m, data = MLM.env, permutations = 1000)
MLM.bray.ord.elev.fit
```



## Vectors in ordination space

```{r NMDS-25,fig.height=5, results='hide'}
plot(MLM.bray.ord, display = "sites", type = "n")
points(MLM.bray.ord,display = "sites", cex = 2)
plot(MLM.bray.ord.elev.fit, add = TRUE)
ordisurf(MLM.bray.ord,MLM.env$elevation_m, add = TRUE)
```


