---
title: "Community data analysis using the vegan package in R"
author: "Naupaka Zimmerman "
date: "March 4, 2016"
output:
  ioslides_presentation:
    fig_height: 6
    fig_width: 7
    highlight: tango
    keep_md: yes
    widescreen: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "", cache = TRUE, crop = TRUE)
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)
```


## Packages installed and loaded?

```{r load-libraries, eval=FALSE}
install.packages("vegan", dependencies = TRUE)
library("vegan")
```


## Loading your OTU table into R

```{r loading-MLM, results='hide'}
MLM.otus <- read.csv("data/MLM_data_otus.csv", row.names = 1, header = TRUE)
MLM.env <- read.csv("data/MLM_data_env.csv", header = TRUE)
```



## Loading your OTU table into R

```{r loading-MLM-2, results='hide'}
MLM.otus <- read.csv("data/MLM_data_otus.csv", row.names = 1, header = TRUE)
MLM.env <- read.csv("data/MLM_data_env.csv", header = TRUE)
```

```{r head-BCI}
head(MLM.otus[,1:3], n = 3)
```



# Basic data summarizing



## Summarizing data with `apply()` | sums and sorting

**Sum of rows**
```{r MLM-apply-1}
sum.of.rows <- apply(MLM.otus, 1, sum)
sort(sum.of.rows, decreasing = TRUE)[1:8] #top 8 rows (plots) 
```



## Summarizing data with `apply()` | sums and sorting

**Sum of rows**
```{r BCI-apply-2}
sum.of.rows <- apply(MLM.otus, 1, sum)
sort(sum.of.rows, decreasing = TRUE)[1:8] #top 8 rows (plots) 
```
**Sum of columns**
```{r BCI-apply-3}
sum.of.columns <- apply(MLM.otus, 2, sum)
sort(sum.of.columns, decreasing = TRUE)[1:3] #top 3 columns (species)
```



## Summarizing data with `apply()` | sums and sorting

**Number of plots in which each spp. occurs**
```{r MLM-apply-4}
spec.pres <- apply(MLM.otus > 0, 2, sum) 
sort(spec.pres, decreasing = TRUE)[1:18]
```



# Data Transformation



## Basic data transformation

**Square root transformation**
```{r MLM-sqrt}
head(MLM.otus[,1:3], n = 3)
```



## Basic data transformation

**Square root transformation**
```{r MLM-sqrt-2}
head(MLM.otus[,1:3], n = 3)
spec.sqrt <- sqrt(MLM.otus)
head(spec.sqrt[,1:3], n = 3)
```



## Data transformation in vegan with `decostand()`

**Total**
```{r MLM-decostand-total}
head(MLM.otus[,1:3], n = 3)
```



## Data transformation in vegan with `decostand()`

**Total**
```{r decostand-total-2}
head(MLM.otus[,1:3], n = 3)
spec.total <- decostand(MLM.otus, method = "total", MARGIN = 1) # by rows (sites)
head(spec.total[,1:3], n = 3)
```



## Data transformation in vegan with `decostand()`

**Maximum**
```{r decostand-max-1}
head(MLM.otus[,1:3], n = 3)
```



## Data transformation in vegan with `decostand()`

**Maximum** 
```{r decostand-max-2}
head(MLM.otus[,1:3], n = 3)
spec.max <- decostand(MLM.otus, method = "max", MARGIN = 2) # by columns (species)
head(spec.max[,1:3], n = 3)
```



## Data transformation in vegan with `decostand()`

**Presence-Absence**
```{r decostand-pa-1}
head(MLM.otus[,1:3], n = 3)
```



## Data transformation in vegan with `decostand()`

**Presence-Absence**
```{r decostand-pa-2}
head(MLM.otus[,1:3], n = 3)
spec.pa <- decostand(MLM.otus, method = "pa")
head(spec.pa[,1:3], n = 3)
```



## Data transformation in vegan with `decostand()`

**Hellinger (Legendre & Gallagher 2001)**    
Square root of method "total"
```{r decostand-hellinger-1}
head(MLM.otus[,1:3], n = 3)
```



## Data transformation in vegan with `decostand()`

**Hellinger (Legendre & Gallagher 2001)**     
Square root of method "total"
```{r decostand-hellinger-2}
head(MLM.otus[,1:3], n = 3)
spec.hellinger <- decostand(MLM.otus, method = "hellinger", MARGIN = 1) # on rows (sites)
head(spec.hellinger[,1:3], n = 3)
```



## Data transformation in vegan with `decostand()`

**Wisconsin double standardization**      
Shortcut function for standardizing species to maximum, then sites by totals.  
```{r wisconsin-1}
head(MLM.otus[,1:3], n = 3)
```



## Data transformation in vegan with `decostand()`

**Wisconsin double standardization**    
Shortcut function for standardizing species to maximum, then sites by totals.
```{r wisconsin-2}
head(MLM.otus[,1:3], n = 3)
spec.wisc <- wisconsin(MLM.otus)
head(spec.wisc[,1:3], n = 3)
```



# Calculating community distances



## Calculating distances with `vegdist()` | so many distance metrics, so little time!

### Many different community distance metrics are available in `vegdist()`    
*manhattan, euclidean, canberra, bray, kulczynski, jaccard, gower, altGower, morisita, horn, mountford, raup, binomial, chao, or cao*



## Calculating distances with `vegdist()` | so many distance metrics, so little time!

### Many different community distance metrics are available in `vegdist()`   
*manhattan, euclidean, canberra, bray, kulczynski, jaccard, gower, altGower, morisita, horn, mountford, raup, binomial, chao, or cao*

```{r dist-jacc-1}
spec.jaccpa <- vegdist(MLM.otus, method = "jaccard", binary = TRUE)
# returns an object of class 'dist'
str(spec.jaccpa) 
```



## Calculating distances with `vegdist()` | so many distance metrics, so little time!

```{r dist-jacc-2}
as.matrix(spec.jaccpa)[1:4,1:4]
```




## Calculating distances with `vegdist()` | so many distance metrics, so little time!

### How to choose a good one for your data set?  
### First step, read the help for vegdist
```{r vegdist-help, eval=FALSE}
?vegdist
```



## Calculating distances with `vegdist()` | so many distance metrics, so little time!

### Second, try `rankindex()`    
Higher rank correlations indicate better separation along gradients
```{r rankindex-1}
rank.elev <- rankindex(MLM.env$elevation_m, MLM.otus, indices = 
              c("bray", "euclid", "manhattan", "horn"), method = "spearman")
rank.elev
```



## Calculating distances with `vegdist()` | so many distance metrics, so little time!

### Second, try `rankindex()`    
Can also use on standardized data
```{r rankindex-2}
rank.elev.wisc <- rankindex(MLM.env$elevation_m, wisconsin(MLM.otus), indices = 
              c("bray", "euclid", "manhattan", "horn"), method = "spearman")
rank.elev.wisc
```



## Calculating distances with `vegdist()` | comparison

```{r rankindex-3}
sort(rank.elev, decreasing = TRUE)
sort(rank.elev.wisc, decreasing = TRUE)
```


# Diversity metrics



## Alpha diversity

Basic counts of richness for each plot or site
```{r alpha-1}
site.richness <- apply(MLM.otus > 0, 1, sum)
site.richness[1:7]
```



## Alpha diversity | Other metrics

Fisher's alpha
```{r alpha-2}
site.fisher <- fisher.alpha(MLM.otus)
site.fisher[1:7]
```



## Alpha diversity | Other metrics

Shannon diversity
```{r alpha-3}
site.shannon <- diversity(MLM.otus, index = "shannon", MARGIN = 1)
site.shannon[1:7]
```



## Rarefaction

This is the same as `apply(BCI > 0, MARGIN = 1, sum)`    
it gives the species count for each plot
```{r rarefac-1}
MLM.S <- specnumber(MLM.otus)
```
This finds the plot with the least number of individuals
```{r rarefac-2}
# could also use rowsums() instead of apply()
MLM.raremax <- min(apply(MLM.otus, 1, sum))
```



## Rarefaction

Rarefy BCI species matrix to the minimum number of individuals in any plot    
and plot the relationship between observed and rarefied counts (plus add 1-1 line)
```{r rarefac-3, fig.height=4}
MLM.Srare <- rarefy(MLM.otus, MLM.raremax)
plot(MLM.S, MLM.Srare, xlab = "Observed No. of Species", ylab = "Rarefied No. of Species")
abline(0, 1)
```



## Rarefaction

Put it all together
```{r rarefac-4, fig.height=5.5}
rarecurve(MLM.otus, step = 20, sample = MLM.raremax, col = "blue", cex = 0.6)
```



## Beta diversity

Multivariate homogeneity of groups dispersions
```{r beta-1}
MLM.bray <- vegdist(MLM.otus, method = "bray")
(MLM.bray.bdisp <- betadisper(MLM.bray,group = as.factor(MLM.env$site_ID)))
```



## Beta diversity {.smaller}

Multivariate homogeneity of groups dispersions
```{r beta-2}
permutest(MLM.bray.bdisp)
```




## Beta diversity

Plot of within-group multivariate dispersion
```{r beta-3, fig.height=4.5}
plot(MLM.bray.bdisp)
```



## Beta diversity

Boxplot of within-group multivariate dispersion
```{r beta-4, fig.height=5.5}
boxplot(MLM.bray.bdisp)
```



## Beta diversity | ANOSIM

Analysis of similarities 
```{r beta-5, fig.height=4.5}
(MLM.bray.anosim <- anosim(MLM.bray, MLM.env$site_ID))
```



## Beta diversity | ANOSIM

Analysis of similarities 
```{r beta-6, fig.height=4.5}
plot(MLM.bray.anosim)
```


## PERMANOVA using `adonis`

Analysis of variance using distance matrices and for fitting linear models to distance matrices
```{r adonis}
adonis(MLM.otus ~ MLM.env$elevation_m)
```

